Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-10-09 morphology_paper.Rmd
moved and deteled some sections from results wave section. spelling checked

@@

Our data show the western side of the peninsula experiences higher significant wave heights and variation in wave heights compared to the eastern side of the peninsula on both an annual and seasonal scale (see figure \ref{fig:wave variable plot}).  Mean significant wave height ranged from 0.5 ± 0.2 m (mean ± SD) at St. Helena Bay to 2.3  ± 0.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side of the peninsula it ranged from 0.9 ± 0.4 m at Buffelsbaai to 1.9 ± 0.6 m at Betty’s Bay (see figure \ref{fig:wave variable plot}A). The maximum significant wave height ranged from 1.4 m at St. Helena Bay to 4.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side ranged from 2.6 m at Buffelsbaai to 4.2 m at Betty’s Bay. Mean peak period for sites on the western side of the peninsula ranged from 10.3 ± 3.5 s at St. Helena Bay to 11.0 ± 2.0 s at Olifantsbos, and ranged from 10.6 ± 3.0 s at Buffelsbaai to 10.8 ± 2.2 s at Betty’s Bay (see figure \ref{fig:wave variable plot}B). The data shows no trend in mean peak period for the coastline. These data show that the western side of the peninsula has a lower variation (SD) compared to the eastern side of the peninsula, with Miller’s point showing the highest variation across timescales. Maximum peak period ranged from 18.9 s at St. Helena Bay to 18.4 s at Olifantsbos for the western side of the peninsula, and ranged from 18.0 s at Buffelsbaai to 18.0 s at Betty’s Bay on the eastern side of the peninsula. When wave variables were grouped by region significant differences were found for both mean significant swell height and mean peak period. The western region The western region experiences a significantly higher mean significant swell height (p < 0.05; see figure \ref{fig:wave variable plot}C) and mean peak period (p < 0.05; see figure \ref{fig:wave variable plot}D). These wave heights and their variability are modulated by the relative sheltering against the predominant south-westerly swell direction and the size of the fetch for local wave generation.

@@

```{r wave rose, echo=FALSE, warning = FALSE, eval= TRUE, fig.align="center",fig.cap='\\label{fig:wave rose}Wave rose figure'}

@@

Data for median wave direction showed no trend across timescales; however the variation (SD) in wave direction is lower on the western side of the peninsula compared to the eastern side both annually and for winter. Summer data were the exception and showed lower variation for the eastern side of the peninsula (see figure \ref{fig:wave rose}). Wave variables divided into Annual (top), Austral summer (middle) and Austral winter (bottom) respectively. This phenomenon can be explained by the sheltering effect of the Cape Peninsula, resulting in a wave climate with a narrower directional spreading on the eastern side of the peninsula (refer to Fig. 7).

@@

There were clear patterns and clines in both the temperature and wave exposure data. The thermal and wave exposure regimes around the Cape Peninsula are driven by a complex interaction between wind, temperature and wave metrics, which do not act independently but instead influence one another. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006], which False Bay is sheltered from [@Shipley1964; @Atkins1970-bx; @Dufois2012-vy]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012-ks; @Leliaert2000a]. It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.
The directional sheltering effect of the Cape Peninsula, against the dominant swell direction is clearly observed in the wave exposures presented in figure \ref{fig:wave power map}. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events along the western side of the peninsula [@Rouault2010-gi; @Field1980a]. In general, the western side of the peninsula is more exposed than the eastern side (False Bay) and experiences significanlty higher significant swell height and significant peak period. This was also reflected in the wave exposure categories which show a trend of decreasing wave exposure (wave power) around the peninsula and into False Bay. There are also differences between the types of waves that each side of the peninsula experiences. The wave analysis shows the western side of the peninsula is not only more exposed but the hydrodynamics are largely swell driven, while on the eastern side most of the wave energy is driven by a combination of swell and wind with the latter being the predominant driver.

>>>>

In general, the mean temperature increases for sites located within False Bay and sites located within False Bay also exhibited less variation around the mean temperature compared to sites on the western side of the peninsula(\ref{fig:Temperature variables plot}A). This is verified by \ref{fig:Temperature variables plot}B, which shows a significantly lower (p<0.05) median temperature and less variation in temperature for sites located on the western side of the peninsula compared to sites on the eastern side.

@@

Our data show the western side of the peninsula experiences higher significant wave heights and variation in wave heights compared to the eastern side of the peninsula on both an annual scale (see figure \ref{fig:wave variable plot}). Mean significant wave height ranged from 0.5 ± 0.2 m (mean ± SD) at St. Helena Bay to 2.3  ± 0.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side of the peninsula it ranged from 0.9 ± 0.4 m at Buffelsbaai to 1.9 ± 0.6 m at Betty’s Bay (see figure \ref{fig:wave variable plot}A). The maximum significant wave height ranged from 1.4 m at St. Helena Bay to 4.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side ranged from 2.6 m at Buffelsbaai to 4.2 m at Betty’s Bay. Mean peak period for sites on the western side of the peninsula ranged from 10.3 ± 3.5 s at St. Helena Bay to 11.0 ± 2.0 s at Olifantsbos, and ranged from 10.6 ± 3.0 s at Buffelsbaai to 10.8 ± 2.2 s at Betty’s Bay (see figure \ref{fig:wave variable plot}B). The data shows no trend in mean peak period for the coastline. These data show that the western side of the peninsula has a lower variation (SD) compared to the eastern side of the peninsula, with Miller’s point showing the highest variation across timescales. Maximum peak period ranged from 18.9 s at St. Helena Bay to 18.4 s at Olifantsbos for the western side of the peninsula, and ranged from 18.0 s at Buffelsbaai to 18.0 s at Betty’s Bay on the eastern side of the peninsula. When wave variables were grouped by region significant differences were found for both mean significant swell height and mean peak period. The western region The western region experiences a significantly higher mean significant swell height (p < 0.05; see figure \ref{fig:wave variable plot}C) and mean peak period (p < 0.05; see figure \ref{fig:wave variable plot}D).

@@

```{r wave rose, echo=FALSE, warning = FALSE, eval= TRUE, fig.align="center",fig.cap='\\label{fig:wave rose}Wave rose figure. Wave variables divided into Annual (top), Austral summer (middle) and Austral winter (bottom) respectively'}

@@

Data for median wave direction showed no trend across timescales; however the variation (SD) in wave direction is lower on the western side of the peninsula compared to the eastern side both annually and for winter. Summer data were the exception and showed lower variation for the eastern side of the peninsula (see figure \ref{fig:wave rose}).

@@

There were clear patterns and clines in both the temperature and wave exposure data. The thermal and wave exposure regimes around the Cape Peninsula are driven by a complex interaction between wind, temperature and wave metrics, which do not act independently but instead influence one another. The hydrodynamic regime is modulated by the relative sheltering against the predominant south-westerly swell direction and the size of the fetch for local wave generation. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006], which False Bay is sheltered from [@Shipley1964; @Atkins1970-bx; @Dufois2012-vy]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012-ks; @Leliaert2000a]. It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.
The directional sheltering effect of the Cape Peninsula, against the dominant swell direction is clearly observed in the wave exposures presented in figure \ref{fig:wave power map}. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events along the western side of the peninsula [@Rouault2010-gi; @Field1980a]. In general, the western side of the peninsula is more exposed than the eastern side (False Bay) and experiences significanlty higher significant swell height and significant peak period. This was also reflected in the wave exposure categories which show a trend of decreasing wave exposure (wave power) around the peninsula and into False Bay. There are also differences between the types of waves that each side of the peninsula experiences. The wave analysis shows the western side of the peninsula is not only more exposed but the hydrodynamics are largely swell driven, while on the eastern side most of the wave energy is driven by a combination of swell and wind with the latter being the predominant driver.
